PEER ASSESSMENT 1 REPRODUCIBLE RESEARCH
========================================================

I will first set echo to TRUE so I won't have to do it each time I create a chunk in r. 

``` {r setoptions, echo = TRUE}
opts_chunk$set(echo = TRUE)

```


Libraries to call:

``` {r}
library(sqldf)
library(lattice)
```
### Loading and Preprocessing the data

I will get the data online, on a zip file. I create a temp file to save the zip, and then unzip it and read it into 'pa1'. Show first 5 rows of pa1 to check. Set field steps to numeric.

```{r getdata, cache=TRUE}
setInternet2(TRUE)
temp <- tempfile()
download.file("https://github.com/andresborrerom/RepData_PeerAssessment1/raw/master/activity.zip",temp)
pa1 <- read.csv(unz(temp, "activity.csv"))
pa1$steps <- as.numeric(pa1$steps)
head(pa1,5)
unlink(temp)
```

### What is mean total number of steps taken per day?

1. Histogram total steps.   
Using SQLDF I will sum steps by unique dates, and then plot a histogram.

```{r plothistogramsteps, fig.height=5, fig.width= 5}
stepsbydate <- sqldf("select date, sum(steps) from pa1 group by date", drv = "SQLite")
stepsbydate[,"sum(steps)"] <- as.numeric(stepsbydate[,"sum(steps)"])
hist( stepsbydate[,"sum(steps)"],7)
```

2. Calculate mean and median

```{r mean_and_median}
promedio <- mean(stepsbydate[,"sum(steps)"], na.rm = TRUE)
mediana <- median(stepsbydate[,"sum(steps)"], na.rm = TRUE)
```

The mean is `r promedio`. 
The median is `r mediana`. 


### What is the average daily activity pattern?

I create a matrix with interval in column 1 and the steps averaged by date in column 2. Then I plot it with x-axis the interval.

```{r fig.height = 4}
meaninterval <- sqldf("select interval, avg(steps) from pa1 group by interval", drv = "SQLite")
plot(meaninterval)
```

``` {r intervalmaxsteps}
m <- subset(meaninterval, meaninterval[,2] == max(meaninterval[,2]))

```

The interval with the maximum steps average is `r m$interval`


### Imputing missing values



``` {r}
p <- subset(pa1, is.na(pa1$steps))

```

1. Total number of rows with NAs: `r nrow(p)`

My strategy for imputting missing values is to group the steps by interval and day of the week (1 through 7). I will imput the mean of these groups into the missing values. 

First I will create the new column weekday (values 1:7): 

``` {r daycolumn, cache = TRUE}
num_day <- matrix(rep(0, nrow(pa1)),nrow(pa1),1)
num_day <- data.frame(num_day)
pa1_w <- cbind(pa1,num_day)
day = 0
for (i in 1:nrow(pa1)){
  if(pa1_w$interval[i] == 0){
    if(day < 7) day = day +1
    else day = 1
  }
  pa1_w$num_day[i] <- day
}
pa1_w[2000:2020,]

```

Now,  I will replace the NAs with the mean of the same interval and weekday, ignoring NAs. Then I will imput the values where missing. 

```{r imputvalues, cache=TRUE}
avg_wd_int <- sqldf("select interval, num_day, avg(steps) from pa1_w group by interval,num_day", drv = "SQLite")

head(avg_wd_int,10)
for (i in 1:nrow(pa1_w)){
  if(is.na(pa1_w$steps[i])){
    temporal <- subset(avg_wd_int, avg_wd_int$num_day == pa1_w$num_day[i] & 
                         avg_wd_int$interval == pa1_w$interval[i])
    pa1_w$steps[i]<-temporal[,"avg(steps)"] 
  }  
}

head(pa1_w,5)
```

(Above: New data set with missing values filled in (head)).

Calculate total steps per day:
```{r plothistogramsteps2, fig.height=5, fig.width= 5}
stepsbydate2 <- sqldf("select date, sum(steps) from pa1_w group by date", drv = "SQLite")
stepsbydate2[,"sum(steps)"] <- as.numeric(stepsbydate2[,"sum(steps)"])
hist( stepsbydate2[,"sum(steps)"],7)
```

2. Calculate mean and median

```{r mean_and_median_2}
promedio2 <- mean(stepsbydate2[,"sum(steps)"], na.rm = FALSE)
mediana2 <- median(stepsbydate2[,"sum(steps)"], na.rm = FALSE)
```

The mean of the new with filled in data is `r promedio2`. 
The median of the new with filled in data is `r mediana2`. 
 
Taking this new calculated data, we have aproximately **`r round(promedio2 - promedio,0)`** more steps in average per day, and a greater median by **`r round(mediana2 - mediana,0)`** steps.

### Are there differences in activity patterns between weekdays and weekends?

I can use my new column created numday to be the division weekday or weekend. Because 2012-10-01 was Monday, and I have it as numday = 1,   weekends will be where numday = 6 or 7. 

```{r weekday_or_weekend, cache = TRUE}
for (i in 1:nrow(pa1_w)){
  if(pa1_w$num_day[i] == 6 | pa1_w$num_day[i] == 7) pa1_w$num_day[i] = "weekend"
  else pa1_w$num_day[i] = "weekday"
}

pa1_w[1438:1445,]
```

Now, create with SQL commands, a data frame with the average number of steps taken per interval when weekday and when weekend.

```{r avg_interval_weekday_weekend, fig.width = 10}
avg_int_wd_we <- sqldf("select num_day, interval, avg(steps) from pa1_w group by interval, num_day", drv     = "SQLite")


head(avg_int_wd_we,5)

head(avg_int_wd_we[,2],5)
xyplot(avg_int_wd_we[,3]~avg_int_wd_we[,2]|avg_int_wd_we[,1], xlab = "intervals", ylab = "average steps")

```